---
title: "Define populations"
author: "Sajal Sthapit"
date: "6/21/2022"
output: html_document
---
# How to define subpopulations
RsB and XP-EHH compare two populations at a time.
In Rsb and xpEHH calculations, Pop1 is used as the reference, and Pop2 is the population in which selection is happening. Therefore, a positive RSB or XPEHH value indicates selection in pop2 wrt reference. Negative values indicate selection in reference population (pop1) wrt to pop2.

3/25/21:
population name convention: habit . population type/name

iHS - run on all populations
Rsb (ratio of EHHS between populations) and xpEHH:
a) run all spring populations with spr.anc as reference and so on for winter
Does (a) give a meaningful answer or new information compared to iHS?
b) split spring and winter into pre61 and pos61 populations and compare selection between the two. For habit/regionals and market classes, split the population into two equal halves and then compare selection before and after. Rsb and xpEHH calculations do not utilize ancestral allele designation information. For state populations, use the previous approach of comparing a state populations with varieties from other regions and not adjacent states.
For methods and results:
Intra population statistic (iHS) was run on all populations. For inter-population statistic (Rsb and xpEHH) we wanted to examine directional selection with progress of plant breeding, i.e., what regions had been selected with respect to older populations. This applies to market classes and regionals. For state populations, we looked at selection in state with respect to varieties from other regions and states.
The drawback of doing different populations for large populations is that if regions are selected in one population, then corresponding regions may be reported as selected in another population just because they are different.

```{r Define subpopulations}
library(tidyverse)
source("functions/functions_for_selection_sweep_analysis.R") # load custom functions for the analysis
samples <- read_tsv("data/variety_details.txt") 

pop <- list()
# ordering the samples first reduces the need to order at the point of selection. However, ordering varieties alters the order of line names. Hence, it is safer to create a new df for pop selection only without altering the original vars
vars <- samples %>% arrange(Year, ACNO)
# both, all spring and winter ----
pop[["both.all"]] <- vars[[1]]
pop$both.all1 <- vars[[1]] %>% first_half
pop$both.all2 <- vars[[1]] %>% second_half
pop$both.eas <- vars %>% filter(Region == "Eastern") %>% .[[1]]
pop$both.eas1 <- pop$both.eas %>% first_half
pop$both.eas2 <- pop$both.eas %>% second_half
pop$both.gpl <- vars %>% filter(Region == "Great Plains") %>% .[[1]]
pop$both.gpl1 <- pop$both.gpl %>% first_half
pop$both.gpl2 <- pop$both.gpl %>% second_half
pop$both.nor <- vars %>% filter(Region == "Northern") %>% .[[1]]
pop$both.nor1 <- pop$both.nor %>% first_half
pop$both.nor2 <- pop$both.nor %>% second_half
pop$both.pac <- vars %>% filter(Region == "Pacific") %>% .[[1]]
pop$both.pac1 <- pop$both.pac %>% first_half
pop$both.pac2 <- pop$both.pac %>% second_half
pop$both.pnw <- vars %>% filter(Region == "PNW") %>% .[[1]]
pop$both.pnw1 <- pop$both.pnw %>% first_half
pop$both.pnw2 <- pop$both.pnw %>% second_half
pop$spr.all <- vars %>% filter(Habit == "spring") %>% .[[1]]
pop$spr.spr1 <- pop$spr.all %>% first_half
pop$spr.spr2 <- pop$spr.all %>% second_half
pop$win.all <- vars %>% filter(Habit == "winter") %>% .[[1]]
pop$win.win1 <- pop$win.all %>% first_half
pop$win.win2 <- pop$win.all %>% second_half

# regional populations ----
pop$spr.gpl <- vars %>% filter(Habit == "spring", Region == "Great Plains") %>% .[[1]]
# spr.gpl only has 26 varieties, so did not split it
pop$spr.nor <- vars %>% filter(Habit == "spring", Region == "Northern") %>% .[[1]]
pop$spr.nor1 <- pop$spr.nor %>% first_half
pop$spr.nor2 <- pop$spr.nor %>% second_half
pop$spr.pac <- vars %>% filter(Habit == "spring", Region == "Pacific")  %>% .[[1]]
pop$spr.pac1 <- pop$spr.pac %>% first_half
pop$spr.pac2 <- pop$spr.pac %>% second_half
pop$spr.pnw <- vars %>% filter(Habit == "spring", Region == "PNW")      %>% .[[1]]
pop$spr.pnw1 <- pop$spr.pnw %>% first_half
pop$spr.pnw2 <- pop$spr.pnw %>% second_half
pop$win.eas <- vars %>% filter(Habit == "winter", Region == "Eastern")  %>% .[[1]]
pop$win.eas1 <- pop$win.eas %>% first_half
pop$win.eas2 <- pop$win.eas %>% second_half
pop$win.gpl <- vars %>% filter(Habit == "winter", Region == "Great Plains") %>% .[[1]]
pop$win.gpl1 <- pop$win.gpl %>% first_half
pop$win.gpl2 <- pop$win.gpl %>% second_half
pop$win.nor <- vars %>% filter(Habit == "winter", Region == "Northern") %>% .[[1]]
# win.nor only had 35 varieties, so did not split it
pop$win.pnw <- vars %>% filter(Habit == "winter", Region == "PNW")      %>% .[[1]]
pop$win.pnw1 <- pop$win.pnw %>% first_half
pop$win.pnw2 <- pop$win.pnw %>% second_half

# Other regional populations ----
pop$both.eas_Other <- vars %>% filter(Region != "Eastern") %>% .[[1]]
pop$both.gpl_Other <- vars %>% filter(Region != "Great Plains") %>% .[[1]]
pop$both.nor_Other <- vars %>% filter(Region != "Northern") %>% .[[1]]
pop$both.pac_Other <- vars %>% filter(Region != "Pacific") %>% .[[1]]
pop$both.pnw_Other <- vars %>% filter(Region != "PNW") %>% .[[1]]
pop$spr.gpl_Other <- vars %>% filter(Habit == "spring", Region != "Great Plains") %>% .[[1]]
pop$spr.nor_Other <- vars %>% filter(Habit == "spring", Region != "Northern") %>% .[[1]]
pop$spr.pac_Other <- vars %>% filter(Habit == "spring", Region != "Pacific") %>% .[[1]]
pop$spr.pnw_Other <- vars %>% filter(Habit == "spring", Region != "PNW") %>% .[[1]]
pop$win.eas_Other <- vars %>% filter(Habit == "winter", Region != "Eastern") %>% .[[1]]
pop$win.gpl_Other <- vars %>% filter(Habit == "winter", Region != "Great Plains") %>% .[[1]]
pop$win.nor_Other <- vars %>% filter(Habit == "winter", Region != "Northern") %>% .[[1]]
pop$win.pnw_Other <- vars %>% filter(Habit == "winter", Region != "PNW") %>% .[[1]]

# Add States with >=20 spring AND winter (both) varieties ----
# filter varieties from each state
states <- map(.x = unique(vars$State), .f = ~ filter(vars, State == .x) %>% .[[1]])
names(states) <- unique(vars$State)
(sel.states <- names(states)[lengths(states) >= 20] %>% sort(.))
# use manual population definition for better code readability
pop$both.CA <- vars %>% filter(State == "California") %>% .[[1]]
pop$both.CO <- vars %>% filter(State == "Colorado") %>% .[[1]]
pop$both.ID <- vars %>% filter(State == "Idaho") %>% .[[1]]
pop$both.IN <- vars %>% filter(State == "Indiana") %>% .[[1]]
pop$both.KS <- vars %>% filter(State == "Kansas") %>% .[[1]]
pop$both.MN <- vars %>% filter(State == "Minnesota") %>% .[[1]]
pop$both.MT <- vars %>% filter(State == "Montana") %>% .[[1]]
pop$both.NE <- vars %>% filter(State == "Nebraska") %>% .[[1]]
pop$both.NY <- vars %>% filter(State == "New York") %>% .[[1]]
pop$both.ND <- vars %>% filter(State == "North Dakota") %>% .[[1]]
pop$both.OR <- vars %>% filter(State == "Oregon") %>% .[[1]]
pop$both.SC <- vars %>% filter(State == "South Carolina") %>% .[[1]]
pop$both.TX <- vars %>% filter(State == "Texas") %>% .[[1]]
pop$both.WA <- vars %>% filter(State == "Washington") %>% .[[1]]

# Manually add other/rest for states based on US production map ----
# exclude.states excludes neighboring states from a different region.
# I could have also filtered by State, but by using Category, I can be explicit about the selection
sel.states[1]
exclude.region <- filter(vars, State == "California") %>% .[["Region"]] %>% unique(.)
exclude.states <- c("Oregon")
pop$both.CA_Other <- filter(vars, Region != exclude.region, !State %in% exclude.states) %>% .[[1]]

sel.states[2]
(exclude.region <- filter(vars, State == "Colorado") %>% .[["Region"]] %>% unique(.))
exclude.states <- c("Arizona", "New Mexico", "Utah", "Wyoming")
pop$both.CO_Other <- filter(vars, Region != exclude.region) %>% .[[1]]

sel.states[3]
(exclude.region <- filter(vars, State == "Idaho") %>% .[["Region"]] %>% unique(.))
exclude.states <- c("Montana", "Nevada", "Wyoming")
pop$both.ID_Other <- filter(vars, Region != exclude.region, !State %in% exclude.states) %>% .[[1]]

sel.states[4]
exclude.region <- filter(vars, State == "Indiana") %>% .[["Region"]] %>% unique(.)
pop$both.IN_Other <- filter(vars, Region != exclude.region) %>% .[[1]]

sel.states[5]
exclude.region <- filter(vars, State == "Kansas") %>% .[["Region"]] %>% unique(.)
exclude.states <- c("Iowa", "Missouri", "Arkansas", "Louisiana")
pop$both.KS_Other <- filter(vars, Region != exclude.region, !State %in% exclude.states) %>% .[[1]]

sel.states[6]
exclude.region <- filter(vars, State == "Minnesota") %>% .[["Region"]] %>% unique(.)
pop$both.MN_Other <- filter(vars, Region != exclude.region) %>% .[[1]]

sel.states[7]
exclude.region <- filter(vars, State == "Montana") %>% .[["Region"]] %>% unique(.)
exclude.states <- c("Idaho")
pop$both.MT_Other <- filter(vars, Region != exclude.region, !State %in% exclude.states) %>% .[[1]]

sel.states[8]
exclude.region <- filter(vars, State == "Nebraska") %>% .[["Region"]] %>% unique(.)
exclude.states <- c("Iowa", "Missouri", "Wyoming")
pop$both.NE_Other <- filter(vars, Region != exclude.region, !State %in% exclude.states) %>% .[[1]]

sel.states[9]
exclude.region <- filter(vars, State == "New York") %>% .[["Region"]] %>% unique(.)
pop$both.NY_Other <- filter(vars, Region != exclude.region) %>% .[[1]]

sel.states[10]
exclude.region <- filter(vars, State == "North Dakota") %>% .[["Region"]] %>% unique(.)
pop$both.ND_Other <- filter(vars, Region != exclude.region) %>% .[[1]]

sel.states[11]
exclude.region <- filter(vars, State == "Oregon") %>% .[["Region"]] %>% unique(.)
exclude.states <- c("California", "Nevada")
pop$both.OR_Other <- filter(vars, Region != exclude.region, !State %in% exclude.states) %>% .[[1]]

sel.states[12]
exclude.region <- filter(vars, State == "South Carolina") %>% .[["Region"]] %>% unique(.)
pop$both.SC_Other <- filter(vars, Region != exclude.region) %>% .[[1]]

sel.states[13]
exclude.region <- filter(vars, State == "Texas") %>% .[["Region"]] %>% unique(.)
pop$both.TX_Other <- filter(vars, Region != exclude.region) %>% .[[1]]

sel.states[14]
exclude.region <- filter(vars, State == "Washington") %>% .[["Region"]] %>% unique(.)
pop$both.WA_Other <- filter(vars, Region != exclude.region) %>% .[[1]]

# Add States with >= 20 spring or winter varieties ----
# Add Category column to vars to apply selection
vars <- mutate(vars, Category = str_c(Habit, State, sep = " "))
rm(states, sel.states) # clear from both
states <- map(.x = unique(vars$Category), .f = ~ filter(vars, Category == .x) %>% .[[1]])
names(states) <- unique(vars$Category)
# Select spring and winter states with >= 20 varieties
lengths(states)
sel.states <- names(states)[lengths(states) >= 20] %>% sort(.) # selection of state names
rm(states)
# sometimes manual addition improves code readability
sel.states
pop$spr.CA <- vars %>% filter(Habit == "spring", State == "California")     %>% .[[1]]
pop$spr.ID <- vars %>% filter(Habit == "spring", State == "Idaho")          %>% .[[1]]
pop$spr.MN <- vars %>% filter(Habit == "spring", State == "Minnesota")      %>% .[[1]]
pop$spr.ND <- vars %>% filter(Habit == "spring", State == "North Dakota")   %>% .[[1]]
pop$spr.WA <- vars %>% filter(Habit == "spring", State == "Washington")     %>% .[[1]]
pop$win.CO <- vars %>% filter(Habit == "winter", State == "Colorado")       %>% .[[1]]
pop$win.ID <- vars %>% filter(Habit == "winter", State == "Idaho")          %>% .[[1]]
pop$win.IN <- vars %>% filter(Habit == "winter", State == "Indiana")        %>% .[[1]]
pop$win.KS <- vars %>% filter(Habit == "winter", State == "Kansas")         %>% .[[1]]
pop$win.NE <- vars %>% filter(Habit == "winter", State == "Nebraska")       %>% .[[1]]
pop$win.NY <- vars %>% filter(Habit == "winter", State == "New York")       %>% .[[1]]
pop$win.OR <- vars %>% filter(Habit == "winter", State == "Oregon")         %>% .[[1]]
pop$win.SC <- vars %>% filter(Habit == "winter", State == "South Carolina") %>% .[[1]]
pop$win.WA <- vars %>% filter(Habit == "winter", State == "Washington")     %>% .[[1]]

# Manually add other/rest for spr/winter states based on US production map ----
# Spring MN should be compared with Spring from other regions
# I could have also filtered by State, but by using Category, I can be explicit about the selection
sel.states[1]
exclude.region <- filter(vars, Category == "spring California") %>% .[["Region"]] %>% unique(.)
exclude.states <- c("Oregon")
pop$spr.CA_Other <- filter(vars, Region != exclude.region, !State %in% exclude.states,
                                 Habit == "spring") %>% .[[1]]
sel.states[2]
exclude.region <- filter(vars, Category == "spring Idaho") %>% .[["Region"]] %>% unique(.)
exclude.states <- c("Montana", "Nevada", "Wyoming")
pop$spr.ID_Other <- filter(vars, Region != exclude.region, !State %in% exclude.states,
                                 Habit == "spring") %>% .[[1]]
sel.states[3]
exclude.region <- filter(vars, Category == "spring Minnesota") %>% .[["Region"]] %>% unique(.)
pop$spr.MN_Other <- filter(vars, Region != exclude.region, Habit == "spring") %>% .[[1]]

sel.states[4]
exclude.region <- filter(vars, Category == "spring North Dakota") %>% .[["Region"]] %>% unique(.)
pop$spr.ND_Other <- filter(vars, Region != exclude.region, Habit == "spring") %>% .[[1]]

sel.states[5]
exclude.region <- filter(vars, Category == "spring Washington") %>% .[["Region"]] %>% unique(.)
pop$spr.WA_Other <- filter(vars, Region != exclude.region, Habit == "spring") %>% .[[1]]

sel.states[6]
exclude.region <- filter(vars, Category == "winter Colorado") %>% .[["Region"]] %>% unique(.)
exclude.states <- c("Wyoming", "Utah", "Arizona")
pop$win.CO_Other <- filter(vars, Region != exclude.region, !State %in% exclude.states,
                                 Habit == "winter") %>% .[[1]]
sel.states[7]
exclude.region <- filter(vars, Category == "winter Idaho") %>% .[["Region"]] %>% unique(.)
exclude.states <- c("Montana", "Nevada", "Wyoming")
pop$win.ID_Other <- filter(vars, Region != exclude.region, !State %in% exclude.states,
                                 Habit == "winter") %>% .[[1]]

sel.states[8]
exclude.region <- filter(vars, Category == "winter Indiana") %>% .[["Region"]] %>% unique(.)
pop$win.IN_Other <- filter(vars, Region != exclude.region, Habit == "winter") %>% .[[1]]

sel.states[9]
exclude.region <- filter(vars, Category == "winter Kansas") %>% .[["Region"]] %>% unique(.)
exclude.states <- c("Iowa", "Missouri", "Arkansas", "Louisiana")
pop$win.KS_Other <- filter(vars, Region != exclude.region, !State %in% exclude.states, 
                                 Habit == "winter") %>% .[[1]]

sel.states[10]
exclude.region <- filter(vars, Category == "winter Nebraska") %>% .[["Region"]] %>% unique(.)
exclude.states <- c("Iowa", "Missouri", "Wyoming")
pop$win.NE_Other <- filter(vars, Region != exclude.region, !State %in% exclude.states,
                                 Habit == "winter") %>% .[[1]]

sel.states[11]
exclude.region <- filter(vars, Category == "winter New York") %>% .[["Region"]] %>% unique(.)
pop$win.NY_Other <- filter(vars, Region != exclude.region, Habit == "winter") %>% .[[1]]

sel.states[12]
exclude.region <- filter(vars, Category == "winter Oregon") %>% .[["Region"]] %>% unique(.)
exclude.states <- c("California", "Nevada")
pop$win.OR_Other <- filter(vars, Region != exclude.region, !State %in% exclude.states,
                                 Habit == "winter") %>% .[[1]]

sel.states[13]
exclude.region <- filter(vars, Category == "winter South Carolina") %>% .[["Region"]] %>% unique(.)
pop$win.SC_Other <- filter(vars, Region != exclude.region, Habit == "winter") %>% .[[1]]

sel.states[14]
exclude.region <- filter(vars, Category == "winter Washington") %>% .[["Region"]] %>% unique(.)
pop$win.WA_Other <- filter(vars, Region != exclude.region, Habit == "winter") %>% .[[1]]

# market classes SKIP ----
pop$spr.HRS <- vars %>% filter(MC == "HRS") %>% .[[1]]
#pop$spr.HWS <- vars %>% filter(MC == "HWS") %>% .[[1]]
pop$spr.SWS <- vars %>% filter(MC == "SWS") %>% .[[1]]
pop$win.HRW <- vars %>% filter(MC == "HRW") %>% .[[1]]
pop$win.SRW <- vars %>% filter(MC == "SRW") %>% .[[1]]
pop$win.SWW <- vars %>% filter(MC == "SWW") %>% .[[1]]
# Other market class populations
pop$spr.HRS_Other <- vars %>% filter(Habit == "spring", MC != "HRS") %>% .[[1]]
pop$spr.SWS_Other <- vars %>% filter(Habit == "spring", MC != "SWS") %>% .[[1]]
pop$win.HRW_Other <- vars %>% filter(Habit == "winter", MC != "HRW") %>% .[[1]]
pop$win.SRW_Other <- vars %>% filter(Habit == "winter", MC != "SRW") %>% .[[1]]
pop$win.SWW_Other <- vars %>% filter(Habit == "winter", MC != "SWW") %>% .[[1]]

# split market classes into halves for Rsb and xpEHH
pop$spr.HRS1 <- first_half(pop$spr.HRS)
pop$spr.HRS2 <- second_half(pop$spr.HRS)
#pop$spr.HWS1 <- first_half(pop$spr.HWS)
#pop$spr.HWS2 <- second_half(pop$spr.HWS)
pop$spr.SWS1 <- first_half(pop$spr.SWS)
pop$spr.SWS2 <- second_half(pop$spr.SWS)
pop$win.HRW1 <- first_half(pop$win.HRW)
pop$win.HRW2 <- second_half(pop$win.HRW)
pop$win.SRW1 <- first_half(pop$win.SRW)
pop$win.SRW2 <- second_half(pop$win.SRW)
pop$win.SWW1 <- first_half(pop$win.SWW)
pop$win.SWW2 <- second_half(pop$win.SWW)

saveRDS(pop, "data/populations.rds")
```

# Last variety in a population
For powerpoint and manuscript
```{r first and last variety of a population}
last.var <- map_df(.x = pop, .f = ~ filter(samples, GS.Sample.ID %in% .x) %>% 
                     arrange(desc(Year)) %>% 
                     .[c("Corrected.Sample.ID", "Year", "MC", "Region", "State")] %>% head(., 1))
last.var <- mutate(last.var, Pop = names(pop)) %>% select(., Pop, everything())
first.var <- map_df(.x = pop, .f = ~ filter(samples, GS.Sample.ID %in% .x) %>% 
                      arrange(Year) %>% .[c("Corrected.Sample.ID", "Year", "MC", "Region", "State")] %>% head(., 1))
first.var <- mutate(first.var, Pop = names(pop)) %>% select(., Pop, everything())
fir.las <- rbind(first.var, last.var) %>% arrange(Pop, Year)
write_csv(fir.las, "output/first_last_varieties.csv")
```

```{context}
getwd()
sessionInfo()
```
