---
title: "Calculations of iHS, Rsb, xpEHH, Fst"
author: "Sajal Sthapit"
date: "6/21/2022"
output: html_document
---
# iHS, RSB and XP-EHH calculations
This part can be put to another place
The scans for RSB and XP-EHH are already in the res object. These metric are for population wise comparisons. Hence, I will first need to define the pair of populations to compare and then pass them to the map2 function to run the calculations.
06/03/2021
Rsb is the natural log of the ratio of integrated EHHS of pop1 over pop2. If there is selection in pop1 then inES(pop1)>inES(pop2) and therefore Rsb value is extreme +ve. Selection in pop2 will be extreme -ve value. Same applies for xpEHH (even stated in Sabeti 2007).
Ignore sign in calc_candidate_regions allows you to select only +ve values, hence my target population should be set as pop1 not pop2 as I had done previously.

```{r}
rm(list = ls())
source("functions/functions_for_selection_sweep_analysis.R") # load custom functions for the analysis
library(tidyverse)
library(rehh)
geno <- list(nuc = read_tsv("data/genotype_nucleotide_format_10391_loci_imputed.txt"),
             ab  = read_tsv("data/genotype_AB_format_10391_loci_imputed.txt"))
```

```{r}
# define pop.pairs in script
names(pop) %>% sort()
# calculate iHS
pop.ihs <- c("both.all", "spr.all", "win.all")
ihs <- map(.x = pop.ihs,
            .f = ~ ihh2ihs(res[[.x]][["scan"]], min_maf = 0, freqbin = 1 ) ) 
# Vignette for rehh (7.6) says to set freqbin = 1 for function ihh2ihs when using unpolarized alleles
names(ihs) <- pop.ihs

pop.pairs <- tibble(pop1 = c("both.all2", "both.CA", "both.CO", "both.eas", "both.eas2", 
                             "both.gpl", "both.gpl2", "both.ID", "both.IN", "both.KS",
                             "both.MN", "both.MT", "both.ND", "both.NE", "both.nor", 
                             "both.nor2", "both.pac", "both.pac2", "both.pnw",
                             "both.pnw2", "both.SC", "both.TX", "both.WA",
                             #spring
                             "spr.all", "spr.CA", "spr.gpl", "spr.ID", "spr.MN", 
                             "spr.ND", "spr.nor", "spr.nor2", "spr.pac",
                             "spr.pac2", "spr.pnw", "spr.pnw2", "spr.spr2", "spr.WA",
                             #winter
                             "win.all", "win.CO", "win.eas", "win.eas2", "win.gpl", 
                             "win.gpl2", "win.ID", "win.IN", "win.KS", "win.NE",
                             "win.nor", "win.NY", "win.OR", "win.pnw", "win.pnw2", 
                             "win.SC", "win.WA", "win.win2"),
                    pop2 = c("both.all1", "both.CA_Other", "both.CO_Other", "both.eas_Other", "both.eas1",
                             "both.gpl_Other", "both.gpl1", "both.ID_Other", "both.IN_Other", "both.KS_Other",
                             "both.MN_Other", "both.MT_Other", "both.ND_Other", "both.NE_Other", "both.nor_Other",
                             "both.nor1", "both.pac_Other", "both.pac1", "both.pnw_Other",
                             "both.pnw1", "both.SC_Other", "both.TX_Other", "both.WA_Other",
                             #spring
                             "win.all", "spr.CA_Other", "spr.gpl_Other", "spr.ID_Other", "spr.MN_Other",
                             "spr.ND_Other", "spr.nor_Other", "spr.nor1", "spr.pac_Other",
                             "spr.pac1", "spr.pnw_Other", "spr.pnw1", "spr.spr1", "spr.WA_Other",
                             #winter
                             "spr.all", "win.CO_Other", "win.eas_Other", "win.eas1", "win.gpl_Other", 
                             "win.gpl1", "win.ID_Other", "win.IN_Other", "win.KS_Other", "win.NE_Other",
                             "win.nor_Other", "win.NY_Other", "win.OR_Other", "win.pnw_Other", "win.pnw1", 
                             "win.SC_Other", "win.WA_Other", "win.win1"))

# Calculate RSB and XP-EHH ----
rsb <- map2(.x = pop.pairs$pop1, .y = pop.pairs$pop2,
            .f = ~ ines2rsb(scan_pop1 = res[[.x]][["scan"]],
                           scan_pop2 = res[[.y]][["scan"]], include_freq = TRUE) )
xpe <- map2(.x = pop.pairs$pop1, .y = pop.pairs$pop2,
            .f = ~ ies2xpehh(scan_pop1 = res[[.x]][["scan"]],
                           scan_pop2 = res[[.y]][["scan"]], include_freq = TRUE) )

names(rsb) <- pop.pairs$pop1
names(xpe) <- pop.pairs$pop1
```

# Calculate Fst, Fst candidate regions, and PIC ----
For Fst calculations we are loading data that are in AA/BB format
```{r load data}
g.10k <- read_tsv("data/genotype_AB_format_10391_loci_imputed.txt")
```

# Calculate Fst and PIC
Total population (allele frequency) used for Fst calculations:
For sub-populations of spring + winter (both): N = 753 (entire panel)
For sub-populations of spring varieties: N = 236 (all spring varieties)
For sub-populations of winter varieties: N = 517 (all winter varieties)
```{r}
names(pop) # unsorted names make it easier to select below
names(pop[c(2:19,22,58:71)])
Fst.sub.both <- calc_Fst(inp = g.10k, save.cols = 6, Total.pop = pop$both.all,
                         Sub.pop = pop[c(2:19,22,58:71)])
names(pop[c(20:21,25:34,86:90)])
Fst.sub.spr  <- calc_Fst(inp = g.10k, save.cols = 6, Total.pop = pop$spr.all,
                         Sub.pop = pop[c(20:21,25:34,86:90)])
names(pop[c(23:24,35:44,91:99)])
Fst.sub.win  <- calc_Fst(inp = g.10k, save.cols = 6, Total.pop = pop$win.all,
                         Sub.pop = pop[c(23:24,35:44,91:99)])
pic.sub <- calc_PIC(inp = g.10k, save.cols = 6, Sub.pop = pop)

fst.sub <- cbind(Fst.sub.both, Fst.sub.spr[7:ncol(Fst.sub.spr)], Fst.sub.win[7:ncol(Fst.sub.win)])
rm(Fst.sub.both, Fst.sub.spr, Fst.sub.win)
```

# START HERE for Candidate Regions
# Calculate candidate regions
Moved the script for calc_candidate_regions back as now I feel more comfortable with the script for iHS, Rsb, and xpEHH
Needs rehh, tidyverse, conv_names_to_labels, and g.pos objects

# Calculate and save candidate regions
On 10/21/20, I checked the RSB candidate regions and compared with the input for calc_candidate_regions and found that only positive RSB values were selected for candidate regions.
My previous output only selected for positive threshold values for candidate regions (ignore_sign = TRUE, default is FALSE, even though the the help says otherwise.) For iHS, I need to ignore sign, but doing that increases candidate regions to 2004 from 661. That means, for iHS I may need to use higher thresholds.
The outputs saved on 10/21/20 include negative IHS candidate regions as well. For RSB and XPEHH, we stuck with positive only as we are only interested in loci selected in the target population and not the reference population. The only change from last version is more candidate regions reported in iHS, which is already higher than Rsb and xpehh. It will also not change the manhattan plots, as I use the raw data for that. But the selection map will need to be updated.
# 02/11/21: The global parameters led to too few candidate regions in some states. To be useful for breeding programs, it would make sense to have a minimum number of candidate regions to explore. Hence, for the state population, we also searched for candidate regions by lowering the threshold in order to select the top 10 candidate regions for each statistic.
# To run this chunk, need to have g.pos and conv_names_to_labels() pre-loaded
# 3/29/21: When I specify maxgap, I lose too many candidate regions. A compromise would be to keep unlimited max gap, but increase window size to 95 percentile of gapsize.

```{r set parameters}
# Set parameters ----
g.pos <- res$g.pos
res$scan_hh.parameters %>% as.character()
res$snp.gap.percentile
res$snp.gap.quantile
param <- c(param, # add to the existing parameters
           list(thres = 2, min.mrk = 6, min.perc = 80, ws = 2.5e6, ovr = 0.2, top = 6))
```

Not all markers are included in the iHS, Rsb, and xpEHH results. Hence, when we combine marker names and cM from g.pos, we end up with more markers than N_MRK value in some cases.

```{r scan for candidate regions with defined parameters}
# select scan_hh results only ----
names(res)
# clean the output by removing populations I do not need ----
# Since I was constantly filtering out populations I do not want to plot in Excel, I decided to filter them here. If you do not want to exclude any populations, ignore the following 2 lines and activate the next 2 lines.
(pop.ihs    <- names(res)[-c(45:57,72:85,100:122)])
(pop.rsbxpe <- names(rsb))

# scan for iHS candidate regions ----
cr.ihs <- map(.x = pop.ihs,
              .f = ~ calc_candidate_regions(res[[.x]][["ihs"]],
                                            threshold = param$thres,
                                            ignore_sign = TRUE, #FALSE means only takes +ve values
                                            window_size = param$ws, overlap = param$ws*param$ovr,
                                            min_n_mrk = param$min.mrk,
                                            min_perc_extr_mrk = param$min.perc) )
# gross sum of candidate regions
map_dbl(.x = cr.ihs, .f = ~ nrow(.x)) %>% sum(.)
names(cr.ihs) <- pop.ihs
cr.ihs$both.all
#TOP filter commented out
#cr.ihs.top <- map(.x = cr.ihs, .f = ~arrange(.x, desc(N_MRK), desc(PERC_EXTR_MRK), desc(MAX_MRK)) %>% .[1:(param$top*2), ])
#cr.ihs.top.df <- conv_names_to_labels(inp = cr.ihs.top, sep = "\\.") %>% filter(!is.na(CHR))
cr.ihs.df <- conv_names_to_labels(inp = cr.ihs, sep = "\\.")
table(cr.ihs.df$Category, cr.ihs.df$CHR)

# Identify candidate regions for Rsb ----
cr.rsb <- map(.x = rsb[pop.rsbxpe], # if you want to scan all pops, use rsb
              .f = ~ calc_candidate_regions(scan = .x, threshold = param$thres,
                                            ignore_sign = FALSE, # for RSB we only want +ve
                                            window_size = param$ws, overlap = param$ws*param$ovr,
                                            min_n_mrk = param$min.mrk,
                                            min_perc_extr_mrk = param$min.perc))
map_dbl(.x = cr.rsb, .f = ~ nrow(.x)) %>% sum(.)
cr.rsb.df <- conv_names_to_labels(inp = cr.rsb, sep = "\\.")
table(cr.rsb.df$Category)

# Identify candidate regions for xpEHH ----
cr.xpe <- map(.x = xpe[pop.rsbxpe], # if you want to scan all pops, use xpe
              .f = ~ calc_candidate_regions(scan = .x, threshold = param$thres,
                                            ignore_sign = FALSE, # for XPEHH we only want +ve
                                            window_size = param$ws, overlap = param$ws*param$ovr,
                                            min_n_mrk = param$min.mrk,
                                            min_perc_extr_mrk = param$min.perc))
map_dbl(.x = cr.xpe, .f = ~ nrow(.x)) %>% sum(.)
cr.xpe.df <- conv_names_to_labels(inp = cr.xpe, sep = "\\.")
table(cr.xpe.df$Category)

# Identify candidate regions for Fst next
```
# Calculate candidate regions for Fst
# Format Fst data into the Rsb results format to use calc_candidate_regions function
All that is needed for calc_candidate_regions is 3 columns: Chr, pos, and statistic. If the Statistic name does not match IHS, RSB, XPEHH, it uses the third column and gives a warning message.
```{r}
fst <- map(.x = names(fst.sub[7:ncol(fst.sub)]),
           .f = ~ data.frame(CHR = fst.sub$Chrom, POSITION = fst.sub$pos,
                             RSB = fst.sub[[.x]]))
names(fst) <- names(fst.sub[7:ncol(fst.sub)])
thres <- map(.x = names(fst.sub)[7:ncol(fst.sub)], .f = ~ quantile(fst.sub[[.x]], .95, na.rm = TRUE) %>% .[[1]])

cr.fst <- map2(.x = fst, .y = thres,
              .f = ~ calc_candidate_regions(scan = .x, threshold = .y, pval = FALSE,
                                            ignore_sign = FALSE, # for RSB we only want +ve
                                            window_size = param$ws, overlap = param$ws*param$ovr,
                                            min_n_mrk = param$min.mrk,
                                            min_perc_extr_mrk = param$min.perc))
map_dbl(.x = cr.fst, .f = ~ nrow(.x)) %>% sum(.)
cr.fst.df <- conv_names_to_labels(inp = cr.fst, sep = "\\.")
table(cr.fst.df$Category)
```

# Add SNPid, cM, and other info to the iHS, Rsb, xpEHH, Fst outputs.
The iHS, Rsb, and xpEHH outputs include CHR and POSITION but no marker names. By adding marker names direclty to these results, it will be easier to identify SNP with highest values in a candidate region to then look for patterns across population to write the discussion.
```{r}
g.pos <- mutate(g.pos, chr.pos = str_c(Chrom, "_", pos))

ihs <- map(.x = names(res[1:45]),
           .f = ~ res[[.x]][["ihs"]][["ihs"]] %>% 
             mutate(chr.pos = str_c(CHR, "_", POSITION)) %>% 
             left_join(., select(g.pos, chr.pos, SNPid, cM), by = "chr.pos" ) %>% 
             select(., -chr.pos))
names(ihs) <- names(res[1:45])
ihs$both.all %>% head()

rsb <- map(.x = rsb,
            .f = ~ mutate(.x, chr.pos = str_c(CHR, "_", POSITION)) %>% 
              left_join(., select(g.pos, chr.pos, SNPid, cM), by = "chr.pos" ) %>% 
             select(., -chr.pos))
head(rsb$both.all2)

xpe <- map(.x = xpe,
            .f = ~ mutate(.x, chr.pos = str_c(CHR, "_", POSITION)) %>% 
              left_join(., select(g.pos, chr.pos, SNPid, cM), by = "chr.pos" ) %>% 
             select(., -chr.pos))
head(xpe$both.all2)

# restore g.pos
g.pos <- select(g.pos, -chr.pos)
```

# Combine and preprocess candidate regions ----
Start this chunk by loading the raw outputs of candidate region search from the chunk above
```{r combine and save candidate regions}
rm(df)

cr.ihs.df <- mutate(cr.ihs.df, Stat = "iHS") 
cr.rsb.df <- mutate(cr.rsb.df, Stat = "Rsb") 
cr.xpe.df <- mutate(cr.xpe.df, Stat = "xpEHH")
cr.fst.df <- mutate(cr.fst.df, Stat = "Fst")

# Merge the ihs, rsb and xpehh candidate regions files ----
names(cr.ihs.df)
df <- rbind(cr.ihs.df, cr.rsb.df, cr.xpe.df, cr.fst.df) %>% 
  mutate(Genome = str_sub(CHR, 2, 2), 
         KB = (END - START)/1000,
         POP = str_c(Pops, " (", Stat, " ", 
                      sprintf("%.2f", round(MAX_MRK, digits = 2)), ", ", 
                      # sprintf is to prevent R from dropping trailing zeroes
                      N_EXTR_MRK, "/", N_MRK, ")")) %>% 
  # Select and order the columns
  select(., Genome, CHR, START, END, KB, POP, Stat, Category, Habit, Pops, 
         N_MRK, N_EXTR_MRK, PERC_EXTR_MRK, MEAN_MRK, MAX_MRK, MEAN_EXTR_MRK) %>% # removed cM, SNP and Name
  arrange(., CHR, START, POP)
# divide physical position by a million
df <- mutate(df, START = START/1e6)
df <- mutate(df, END = END/1e6)
# add lines to add Genome (str_sub) and Percentage Extremal Markers.
table(df$Category)
table(df$CHR)
median(df$KB)
group_by(df, Stat) %>% 
  summarise(mean = mean(KB), median = median(KB), min = min(KB), max = max(KB))

cr.param <- str_c("cr-thres", param$thres, 
                  "-mrk", param$min.mrk, 
                  "-per", param$min.perc,
                  "-ws", param$ws/1000, "Kb",
                  "-ovr", param$ovr*100, "per")
scan_hh.param <- paste0("-pol-",  res$scan_hh.parameters$polarized.value, # paste0 converts NA to character
                        "-sgap-", res$scan_hh.parameters$scale.gap.value,
                        "-mgap-", res$scan_hh.parameters$max.gap.value,
                        "-discardBorder-", res$scan_hh.parameters$discard.integration.at.border, "-")
raw.results <- list(ihs = ihs, rsb = rsb, xpe = xpe, fst = fst, 
                    cr.consolidated = df, pic.sub = pic.sub)
saveRDS(raw.results, str_c("rehh-files/images/raw_results_for_cr_search_", Sys.Date(), ".RDS"))
write_csv(df, str_c("rehh-files/images/", cr.param, scan_hh.param, Sys.Date(), ".csv"))
save.image(file = "rehh_completed_analysis.RData")
```

# Identify allelic state for every variety
# By using nucleotide data, I can even say which SNP is selected where
```{r}
load(file = "rehh_completed_analysis.RData")
trans.g <- g.poly[7:ncol(g.poly)] %>% t %>% as_tibble()
names(trans.g) <- g.poly$SNPid
samples <- readRDS("images/samplesdetailsforplotting-2020-08-11.rds")
sum(samples$GS.Sample.ID == names(g.poly[7:ncol(g.poly)])) # sample name orders match
samples <- cbind(samples, trans.g)
write_tsv(samples, "rehh_completed_for_plotting.txt")
# for discussion, simply take the marker in the candidate region, filter the population using the marker and draw cross tabs.
# for example

# FUNCTION DEFINITIONS ----
# This function extracts the marker with the most extreme iHS, Rsb, xpEHH, or Fst value from each candidate region. The list output generated can then be used as input to generate the custom bar plots.
extract_best_markers <- function(cr.df, scan.res, stat.label){ 
  # cr.df = data.frame of candidate regions with CHR, START, END, and Category columns.
  # scan.res = results of iHS, Rsb, xpEHH, or Fst calculation respectively.
  # stat.label = name of the statistics column in the scan.res (IHS, RSB, or XPEHH). Note that the Fst values for fst.scan.res is also stored in the column names RSB to be compatible for the rehh::calc_candidate_regions function.
  pops.with.res <- intersect(names(scan.res), cr.df$Category) # populations that returned candidate regions
  inp <- filter(cr.df, Category %in% pops.with.res) # why are some candidate regions not in the ihs object? The should match.
  marker <- list()
  marker$SNPid <- map(.x = 1:nrow(inp), 
      .f = ~ scan.res[inp$Category][[1]] %>% # isolates the right data.frame with the population-specific scan results
        filter(CHR == inp$CHR[[.x]], POSITION >= inp$START[[.x]], POSITION <= inp$END[[.x]]) %>% 
        arrange(desc(abs(.data[[stat.label]]))) %>% 
        .[["SNPid"]] %>% .[[1]]    ) # works
  marker$details <- map(.x = 1:nrow(inp), 
      .f = ~ scan.res[inp$Category][[1]] %>% # isolates the right data.frame with the population-specific scan results
        filter(CHR == inp$CHR[[.x]], POSITION >= inp$START[[.x]], POSITION <= inp$END[[.x]]) %>% 
        arrange(desc(abs(.data[[stat.label]]))) %>% 
        slice_head(n = 1) %>% 
        mutate(details = str_c(SNPid, " at ", CHR, ":", POSITION/1e6, " MBp")) %>% 
          .[["details"]]    ) # works
 return(marker) 
}

custom_box_plot <- function(inp, x.axis.label = NULL){
  # inp is a tibble that includes a column for y-axis variable (Year), x-axis variable (selected marker), and category variable columns for faceting
  # example inp: inp <- select(samples, marker$SNPid[[1]], Year, Habit, Region, MC)
  x.var <- names(inp[1])
  y.var <- names(inp[2])
  
  p.habit <- ggplot(inp, aes_string(x = x.var, y = y.var)) +
    geom_violin() + geom_boxplot(alpha = 0.5) +
    facet_grid(. ~ Habit) +
    stat_summary(fun = mean, geom = "point", shape = 15, size = 3, color = "black") +
    theme_bw() +
    labs(x = NULL, subtitle = "Habit")
  p.region <- filter(inp, Region != "AK") %>% # Remove Alaska as it only has 4 varieties
    ggplot(., aes_string(x = x.var, y = y.var)) +
    geom_violin() + geom_boxplot(alpha = 0.5) +
    facet_grid(. ~ Region) +
    stat_summary(fun = mean, geom = "point", shape = 15, size = 3, color = "black") +
    theme_bw() +
    labs(x = NULL, subtitle = "Region")
  p.mc <- filter(inp, !MC %in% c("HWS", "HWW", "xxW")) %>% # remove market classes with few varieties
    ggplot(., aes_string(x = x.var, y = y.var)) +
    geom_violin() + geom_boxplot(alpha = 0.5) +
    facet_grid(. ~ MC) +
    stat_summary(fun = mean, geom = "point", shape = 15, size = 3, color = "black") +
    theme_bw() +
    labs(x = NULL, subtitle = "Class")
  
  p.plots <- plot_grid(p.habit, p.region, p.mc, nrow = 1, rel_widths = c(.5, 1, 1))
  p.label <- ggdraw() + draw_label(str_c("Alleles for ", x.axis.label))
  return(plot_grid(p.label, p.plots, ncol = 1, rel_heights = c(0.1, 1)))
}

custom_bar_plot <- function(inp, x.axis.label = NULL){
  # inp is a tibble with 3 columns: first column is the category of variety, second column is the allelic state for the marker (column name), third column is the count n: e.g., Habit | IWB653314 | n. This table is generated by table(Habit, IWB653314) %>% as_tibble()
  # Extract category and marker names
  category <- names(inp[1])
  alleles <- names(inp[2])
  count <- names(inp[3])
  
  # Plots
  out <-  ggplot(inp, aes_string(x = alleles, y = count)) +  
    geom_bar(stat = "identity") +
    facet_grid(reformulate(category)) + # use reformulate to pass the string in category as ~ variable
    theme_bw() +
    labs(x = x.axis.label, y = "Count", subtitle = category)
return(out)

}

draw_combined_custom_plots <- function(inp, marker.snpid, marker.details){
  # inp is a genotype data frame with category columns (Habit, Region, etc.) and all the markers as columns
  # marker.snpid is the SNPid to be plotted
  # marker.details is the descriptive label to use for the plot (e.g., "IWB61100 at 1A:482 MBp")
  out.box <- select(inp, marker.snpid, Year, Habit, Region, MC) %>% custom_box_plot(., marker.details)
  # combine bar plots
  p.bar <- list()
  p.bar$habit <- table("Habit" = inp$Habit, "alleles" = inp[[marker.snpid]]) %>% as_tibble %>% custom_bar_plot()
  p.bar$region <- table("Region" = inp$Region, "alleles" = inp[[marker.snpid]]) %>% as_tibble %>% 
    filter(Region != "AK") %>%  custom_bar_plot()
  p.bar$mc <- table("Class" = inp$MC, "alleles" = inp[[marker.snpid]]) %>% as_tibble %>% 
    filter(!Class %in% c("HWS", "HWW", "xxW")) %>%  custom_bar_plot()
  p.bar$bp <- table("Period" = inp$BP, "alleles" = inp[[marker.snpid]]) %>% as_tibble %>% custom_bar_plot()
  p.bar$decade <- table("Decade" = inp$Decade, "alleles" = inp[[marker.snpid]]) %>% as_tibble %>% custom_bar_plot()
  p.bar$row1 <- plot_grid(p.bar$habit, p.bar$region, p.bar$mc, nrow = 1, rel_widths = c(.4, 1, 1))
  p.bar$row2 <- plot_grid(p.bar$bp, p.bar$decade, nrow = 1, rel_widths = c(.3, 1))
  out.bar <- plot_grid(p.bar$row1, p.bar$row2)
  return(plot_grid(out.box, out.bar, ncol = 1))
}

# APPLY functions ----
marker <- extract_best_markers(cr.df = cr.rsb.df, scan.res = rsb, stat.label = "RSB")

draw_combined_custom_plots(samples, marker$SNPid[[1]], marker$details[[1]])
map(.x = 1:3,
    .f = ~ draw_combined_custom_plots(inp = samples, marker$SNPid[[.x]], marker$details[[.x]]))
```

# Manhattan plots were here
# END of the script