---
title: "Calculations of Rsb and xpEHH"
author: "Sajal Sthapit"
date: "6/21/2022"
output: html_document
---
# RSB and XP-EHH calculations
This part can be put to another place
The scans for RSB and XP-EHH are already in the res object. These metric are for population wise comparisons. Hence, I will first need to define the pair of populations to compare and then pass them to the map2 function to run the calculations.
06/03/2021
Rsb is the natural log of the ratio of integrated EHHS of pop1 over pop2. If there is selection in pop1 then inES(pop1)>inES(pop2) and therefore Rsb value is extreme +ve. Selection in pop2 will be extreme -ve value. Same applies for xpEHH (even stated in Sabeti 2007). Therefore, it is important to note what were assigned as population 1 and 2.
Ignore sign in calc_candidate_regions allows you to select only +ve values, hence my target population should be set as pop1 not pop2 as I had done previously.

```{r message=FALSE, warning=FALSE}
rm(list = ls())
source("functions/functions_for_selection_sweep_analysis.R") # load custom functions for the analysis
library(tidyverse)
library(rehh)
library(pegas) # for Fst calculation
geno <- list(nuc = read_tsv("data/genotype_nucleotide_format_10391_loci_imputed.txt"),
             ab  = read_tsv("data/genotype_AB_format_10391_loci_imputed.txt"))
res <- readRDS("output/scan_hh_ihs_results_polFALSE_sgap2.5MB_mgapNAMB_discardBorderTRUE.rds")
```
# Define population pairs and calculate Rsb, xpEHH, and Fst
```{r message=FALSE, warning=FALSE}
# define pop.pairs in script. Target population for selection is in population 1.
pop.pairs <- tibble(pop1 = c("both.all2", "both.CA", "both.CO", "both.eas", "both.eas2", 
                             "both.gpl", "both.gpl2", "both.ID", "both.IN", "both.KS",
                             "both.MN", "both.MT", "both.ND", "both.NE", "both.nor", 
                             "both.nor2", "both.pac", "both.pac2", "both.pnw",
                             "both.pnw2", "both.SC", "both.TX", "both.WA",
                             #spring
                             "spr.all", "spr.CA", "spr.gpl", "spr.ID", "spr.MN", 
                             "spr.ND", "spr.nor", "spr.nor2", "spr.pac",
                             "spr.pac2", "spr.pnw", "spr.pnw2", "spr.spr2", "spr.WA",
                             #winter
                             "win.all", "win.CO", "win.eas", "win.eas2", "win.gpl", 
                             "win.gpl2", "win.ID", "win.IN", "win.KS", "win.NE",
                             "win.nor", "win.NY", "win.OR", "win.pnw", "win.pnw2", 
                             "win.SC", "win.WA", "win.win2"),
                    pop2 = c("both.all1", "both.CA_Other", "both.CO_Other", "both.eas_Other", "both.eas1",
                             "both.gpl_Other", "both.gpl1", "both.ID_Other", "both.IN_Other", "both.KS_Other",
                             "both.MN_Other", "both.MT_Other", "both.ND_Other", "both.NE_Other", "both.nor_Other",
                             "both.nor1", "both.pac_Other", "both.pac1", "both.pnw_Other",
                             "both.pnw1", "both.SC_Other", "both.TX_Other", "both.WA_Other",
                             #spring
                             "win.all", "spr.CA_Other", "spr.gpl_Other", "spr.ID_Other", "spr.MN_Other",
                             "spr.ND_Other", "spr.nor_Other", "spr.nor1", "spr.pac_Other",
                             "spr.pac1", "spr.pnw_Other", "spr.pnw1", "spr.spr1", "spr.WA_Other",
                             #winter
                             "spr.all", "win.CO_Other", "win.eas_Other", "win.eas1", "win.gpl_Other", 
                             "win.gpl1", "win.ID_Other", "win.IN_Other", "win.KS_Other", "win.NE_Other",
                             "win.nor_Other", "win.NY_Other", "win.OR_Other", "win.pnw_Other", "win.pnw1", 
                             "win.SC_Other", "win.WA_Other", "win.win1"))

# Calculate RSB, XP-EHH, and Fst ----
rsb <- map2(.x = pop.pairs$pop1, .y = pop.pairs$pop2,
            .f = ~ ines2rsb(scan_pop1 = res[[.x]][["scan"]],
                           scan_pop2 = res[[.y]][["scan"]], include_freq = TRUE) )
xpe <- map2(.x = pop.pairs$pop1, .y = pop.pairs$pop2,
            .f = ~ ies2xpehh(scan_pop1 = res[[.x]][["scan"]],
                           scan_pop2 = res[[.y]][["scan"]], include_freq = TRUE) )

names(rsb) <- pop.pairs$pop1
names(xpe) <- pop.pairs$pop1

#Fst calculations
geno$diploid <- geno$nuc[7:ncol(geno$nuc)]
geno$diploid[geno$diploid == "A"] <- "A/A"
geno$diploid[geno$diploid == "C"] <- "C/C"
geno$diploid[geno$diploid == "G"] <- "G/G"
geno$diploid[geno$diploid == "T"] <- "T/T"
colnames(geno$diploid) <- colnames(geno$nuc[7:ncol(geno$nuc)]) # restore column names

#fst calculation
fst <- map2(.x = pop.pairs$pop1, .y = pop.pairs$pop2,
            .f = ~ calc_fst(pop1 = .x, pop2 = .y))
names(fst) <- pop.pairs$pop1
```
# START HERE for Candidate Regions
# Calculate candidate regions
Moved the script for calc_candidate_regions back as now I feel more comfortable with the script for iHS, Rsb, and xpEHH
Needs rehh, tidyverse, conv_names_to_labels, and g.pos objects

# Calculate and save candidate regions
On 10/21/20, I checked the RSB candidate regions and compared with the input for calc_candidate_regions and found that only positive RSB values were selected for candidate regions.
My previous output only selected for positive threshold values for candidate regions (ignore_sign = TRUE, default is FALSE, even though the the help says otherwise.) For iHS, I need to ignore sign, but doing that increases candidate regions to 2004 from 661. That means, for iHS I may need to use higher thresholds.
The outputs saved on 10/21/20 include negative IHS candidate regions as well. For RSB and XPEHH, we stuck with positive only as we are only interested in loci selected in the target population and not the reference population. The only change from last version is more candidate regions reported in iHS, which is already higher than Rsb and xpehh. It will also not change the manhattan plots, as I use the raw data for that. But the selection map will need to be updated.
# 02/11/21: The global parameters led to too few candidate regions in some states. To be useful for breeding programs, it would make sense to have a minimum number of candidate regions to explore. Hence, for the state population, we also searched for candidate regions by lowering the threshold in order to select the top 10 candidate regions for each statistic.
# To run this chunk, need to have g.pos and conv_names_to_labels() pre-loaded
# 3/29/21: When I specify maxgap, I lose too many candidate regions. A compromise would be to keep unlimited max gap, but increase window size to 95 percentile of gapsize.

```{r set parameters}
# Set parameters ----
res$scan_hh.parameters %>% as.character()
res$snp.gap.percentile
res$snp.gap.quantile
param <- list(thres = 2, min.mrk = 6, min.perc = 80, ws = 2.5e6, ovr = 0.2)
```

```{r calculate allele frequencies}
allele.freq <- map(.x = names(res$pop), .f = ~ calc_allele_freq(pop.name = .x))
names(allele.freq) <- names(res$pop)
```
The rehh::calc_candidate_regions() outputs a data frame with candidate regions for each population. Then chunk below combines the different population-specific data frames into one dataframe.

```{r scan for candidate regions with defined parameters}
# select scan_hh results only ----
names(res)
# clean the output by removing populations I do not need ----
# Since I was constantly filtering out populations I do not want to plot in Excel, I decided to filter them here. If you do not want to exclude any populations, ignore the following 2 lines and activate the next 2 lines.
(pop.rsbxpe <- names(rsb))

# Identify candidate regions for Rsb ----
cr.rsb <- map(.x = rsb[pop.rsbxpe], # if you want to scan all pops, use rsb
              .f = ~ calc_candidate_regions(scan = .x, threshold = param$thres,
                                              ignore_sign = TRUE, # for RSB we only want +ve
                                            window_size = param$ws, overlap = param$ws*param$ovr,
                                            min_n_mrk = param$min.mrk,
                                            min_perc_extr_mrk = param$min.perc))
rsb$both.all2 %>% nrow()
map_dbl(.x = cr.rsb, .f = ~ nrow(.x)) %>% sum(.) # gross number of Rsb candidate regions
cr.rsb.df <- conv_names_to_labels(inp = cr.rsb, sep = "\\.") # this throws up a warning even though xpe does not
table(cr.rsb.df$Category)

# Identify candidate regions for xpEHH ----
cr.xpe <- map(.x = xpe[pop.rsbxpe], # if you want to scan all pops, use xpe
              .f = ~ calc_candidate_regions(scan = .x, threshold = param$thres,
                                            ignore_sign = TRUE, # FALSE if you only want +ve
                                            window_size = param$ws, overlap = param$ws*param$ovr,
                                            min_n_mrk = param$min.mrk,
                                            min_perc_extr_mrk = param$min.perc))
map_dbl(.x = cr.xpe, .f = ~ nrow(.x)) %>% sum(.) # = gross number of xpEHH candidate regions
cr.xpe.df <- conv_names_to_labels(inp = cr.xpe, sep = "\\.")
table(cr.xpe.df$Category)

# Identify candidate regions for Fst ----
# Fst filtered for the 5% outliers for each pair wise comparison
quantile.threshold <- map_dbl(.x = names(fst), .f = ~ quantile(fst[[.x]][["FST"]], 0.975, na.rm = TRUE))
names(quantile.threshold) <- names(fst)
cr.fst <- map(.x = names(fst), # if you want to scan all pops, use rsb
              .f = ~ calc_candidate_regions(scan = data.frame(CHR = fst[[.x]][["CHR"]],
                                                              POSITION = fst[[.x]][["POSITION"]],
                                                              RSB = fst[[.x]][["FST"]]), 
                                            # the input for calc_candidate_regions needs to be a data.frame and needs to have a column named IHS, RSB, or XPEHH. If such a column is not found, it uses the third column, but generates warning messages in the console. Hence, I just renamed the column to RSB within this map function to avoid the warning messages.
                                            threshold = quantile.threshold[[.x]],
                                            ignore_sign = TRUE, 
                                            window_size = param$ws, overlap = param$ws*param$ovr,
                                            min_n_mrk = param$min.mrk,
                                            min_perc_extr_mrk = param$min.perc))
names(cr.fst) <- names(fst)
map_dbl(.x = cr.fst, .f = ~nrow(.x)) %>% sum(.)
cr.fst.df <- conv_names_to_labels(inp = cr.fst, sep = "\\.")
```

# Add SNPid, cM, other info, allele frequencies to Rsb, xpEHH, Fst outputs.
The Rsb, and xpEHH outputs include CHR and POSITION but no marker names. By adding marker names directly to these results, it will be easier to identify SNP with highest values in a candidate region to then look for patterns across population to write the discussion.
```{r}
g.pos <- mutate(geno$ab[1:6], chr.pos = str_c(Chrom, "_", pos)) %>% select(., SNPid, cM, chr.pos)

# update Rsb with chr and position
rsb.id <- map(.x = rsb,
            .f = ~ mutate(.x, chr.pos = str_c(CHR, "_", POSITION)) %>% 
              left_join(select(g.pos, chr.pos, SNPid, cM), ., by = "chr.pos" ) %>% 
              select(., -chr.pos) %>% 
              select(., SNPid, cM, RSB, LOGPVALUE, everything()) )

# update Xpe with chr and position
xpe.id <- map(.x = xpe,
            .f = ~ mutate(.x, chr.pos = str_c(CHR, "_", POSITION)) %>% 
              left_join(select(g.pos, chr.pos, SNPid, cM), ., by = "chr.pos" ) %>%
              select(., -chr.pos) %>% 
              select(., SNPid, cM, XPEHH, LOGPVALUE, everything()) )
head(xpe$both.all2)
head(xpe.id$both.all2)

# update fst with chr and position
fst.id <- map(.x = fst,
            .f = ~ mutate(.x, chr.pos = str_c(CHR, "_", POSITION)) %>% 
              left_join(select(g.pos, chr.pos, SNPid, cM), ., by = "chr.pos" ) %>%
              select(., -chr.pos) %>% 
              select(., SNPid, cM, FST, everything()) )

# consolidate all results ----
final <- map2(.x = pop.pairs$pop1, .y = pop.pairs$pop2,
            .f = ~ consolidate_results(pop1 = .x, pop2 = .y) )
names(final) <- pop.pairs$pop1
```

# Combine and preprocess candidate regions ----
Start this chunk by loading the raw outputs of candidate region search from the chunk above
```{r combine and save candidate regions}
rm(df)

cr.rsb.df <- mutate(cr.rsb.df, Stat = "Rsb") 
cr.xpe.df <- mutate(cr.xpe.df, Stat = "xpEHH")
cr.fst.df <- mutate(cr.fst.df, Stat = "Fst")

# Merge the fst, rsb and xpehh candidate regions files ----
names(cr.rsb.df)
df <- rbind(cr.rsb.df, cr.xpe.df, cr.fst.df) %>% 
  mutate(Genome = str_sub(CHR, 2, 2), 
         KB = (END - START)/1000,
         POP = str_c(Pops, " (", Stat, " ", 
                      sprintf("%.2f", round(MAX_MRK, digits = 2)), ", ", 
                      # sprintf is to prevent R from dropping trailing zeroes
                      N_EXTR_MRK, "/", N_MRK, ")")) %>% 
  # Select and order the columns
  select(., Genome, CHR, START, END, KB, POP, Stat, Category, Habit, Pops, 
         N_MRK, N_EXTR_MRK, PERC_EXTR_MRK, MEAN_MRK, MAX_MRK, MEAN_EXTR_MRK) %>% # removed cM, SNP and Name
  arrange(., CHR, START, POP)
# divide physical position by a million
df <- mutate(df, START = START/1e6)
df <- mutate(df, END = END/1e6)
# add lines to add Genome (str_sub) and Percentage Extremal Markers.
table(df$Category)
table(df$CHR)
median(df$KB)
group_by(df, Stat) %>% 
  summarise(mean = mean(KB), median = median(KB), min = min(KB), max = max(KB))

cr.param <- str_c("cr-thres", param$thres, 
                  "-mrk", param$min.mrk, 
                  "-per", param$min.perc,
                  "-ws", param$ws/1000, "Kb",
                  "-ovr", param$ovr*100, "per")
scan_hh.param <- paste0("-pol-",  res$scan_hh.parameters$polarized.value, # paste0 converts NA to character
                        "-sgap-", res$scan_hh.parameters$scale.gap.value,
                        "-mgap-", res$scan_hh.parameters$max.gap.value,
                        "-discardBorder-", res$scan_hh.parameters$discard.integration.at.border, "-")
raw.results <- list(rsb = rsb, xpe = xpe, fst = fst, cr.consolidated = df)
saveRDS(raw.results, "output/raw_results_for_cr_search.RDS")
write_csv(df, str_c("output/cr_consolidated.csv"))
save.image(file = "output/rehh_completed_analysis.RData")
```

# Export SNPids within the candidate regions to text files
The mean_extr_mrk column in the calc_candidate_regions output include mean of absolute values. However, we want to know the sign (+/-) of the statistics in order to ascertain if the selection is happening in population 1 or 2. The chunk below adds the mean_extr_mrk with sign
```{r Export markers in the candidate regions}
# candidate region details are CHR, START, END, Category
ext.rsb <- map(.x = 1:nrow(cr.rsb.df),
                  .f = ~ rsb.id[[cr.rsb.df[.x,][["Category"]]]] %>% 
                    filter(., abs(RSB) >= 2.0, 
                           CHR == cr.rsb.df[.x,][["CHR"]], 
                           POSITION >= cr.rsb.df[.x,][["START"]], 
                           POSITION <= cr.rsb.df[.x,][["END"]]))

ext.xpe <- map(.x = 1:nrow(cr.xpe.df),
                  .f = ~ xpe.id[[cr.xpe.df[.x,][["Category"]]]] %>% 
                    filter(., abs(XPEHH) >= 2.0, 
                           CHR == cr.xpe.df[.x,][["CHR"]], 
                           POSITION >= cr.xpe.df[.x,][["START"]], 
                           POSITION <= cr.xpe.df[.x,][["END"]]))

# use ext.rsb to add mean_extr_mrk_sign column to cr.rsb.df
extr.mrk.stats <- list(rsb.mean = map_dbl(.x = 1:nrow(cr.rsb.df), .f = ~ ext.rsb[[.x]][["RSB"]] %>% mean()),
                       rsb.num  = map_dbl(.x = 1:nrow(cr.rsb.df), .f = ~ ext.rsb[[.x]] %>% nrow()),
                       xpe.mean = map_dbl(.x = 1:nrow(cr.xpe.df), .f = ~ ext.xpe[[.x]][["XPEHH"]] %>% mean()),
                       xpe.num  = map_dbl(.x = 1:nrow(cr.xpe.df), .f = ~ ext.xpe[[.x]] %>% nrow()) )
cr.rsb.df <- mutate(cr.rsb.df, nonABS_mean_extr_mrk = extr.mrk.stats$rsb.mean, stat_sign = if_else(nonABS_mean_extr_mrk > 0, "pos", "neg"))
cr.rsb.df$MEAN_EXTR_MRK - abs(cr.rsb.df$nonABS_mean_extr_mrk) # check to see the results are identical
cr.xpe.df <- mutate(cr.xpe.df, nonABS_mean_extr_mrk = extr.mrk.stats$xpe.mean, stat_sign = if_else(nonABS_mean_extr_mrk > 0, "pos", "neg"))
cr.xpe.df$MEAN_EXTR_MRK - abs(cr.xpe.df$nonABS_mean_extr_mrk) # check to see the results are identical

# updates the names of the ext.rsb and ext.xpe list items indicating whether the stat value is +ve or -ve
names(ext.rsb) <- str_c(cr.rsb.df$stat_sign, cr.rsb.df$CHR, cr.rsb.df$Category, cr.rsb.df$START/1e6, cr.rsb.df$END/1e6, sep = "_")
names(ext.xpe) <- str_c(cr.xpe.df$stat_sign, cr.xpe.df$CHR, cr.xpe.df$Category, cr.xpe.df$START/1e6, cr.xpe.df$END/1e6, sep = "_")

# copy to the right folders
dir.name <- "rsb"
dir.create(dir.name)
map_chr(.x = names(ext.rsb), .f = ~ dir.create(str_c(dir.name, "/", .x)) ) # by using map_chr, the warnings will be in one line instead of multiple lines.
list.of.file.name.vectors <- map(.x = ext.rsb, .f = ~ str_c("./images/", .x[["SNPid"]], ".png")) # create file name vectors to use for copying next
for(i in 1:length(list.of.file.name.vectors)){
  file.copy(from = list.of.file.name.vectors[[i]], # double square brackets to get to the vector of characters instead of a list
            to   = str_c("./", dir.name, "/", names(list.of.file.name.vectors[i]) ) )
  }
# repeat for xpehh
rm(dir.name)
dir.name <- "xpe"
dir.create(dir.name)
map_chr(.x = names(ext.rsb), .f = ~ dir.create(str_c(dir.name, "/", .x)) ) # by using map_chr, the warnings will be in one line instead of multiple lines.
list.of.file.name.vectors <- map(.x = ext.rsb, .f = ~ str_c("./images/", .x[["SNPid"]], ".png")) # create file name vectors to use for copying next
for(i in 1:length(list.of.file.name.vectors)){
  file.copy(from = list.of.file.name.vectors[[i]], # double square brackets to get to the vector of characters instead of a list
            to   = str_c("./", dir.name, "/", names(list.of.file.name.vectors[i]) ) )
  }
```


# Manhattan plots were here
# END of the script
```{r manhattan plots}
manhattanplot(rsb$win.win2)

fst.mrk <- map(.x = pop.pairs$pop1, .f = ~ extract_most_extreme_fst_snp(pop1 = .x) )
names(fst.mkr) <- pop.pairs$pop1
rsb.mrk <- map(.x = pop.pairs$pop1, .f = ~ extract_most_extreme_rsb_snp(pop1 = .x) )
names(rsb.mrk) <- pop.pairs$pop1
xpe.mrk <- map(.x = pop.pairs$pop1, .f = ~ extract_most_extreme_xpehh_snp(pop1 = .x))
names(xpe.mrk) <- pop.pairs$pop1

manhattanplot(fst.id$both.all2, threshold = quantile.threshold[["both.all2"]], cr = cr.fst$both.all2, mrk = fst.mrk[["both.all2"]])

# write the manhattan plots to a pdf file
pdf(file = "output/manhattanplots.pdf", width = 10, height = 8)
map(.x = pop.pairs$pop1, .f = ~ custom_manhattan(pop1 = .x))
dev.off()




```